A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast

### 2021-11-04 analysis/paper/paper.Rmd
Analysis of individual sites

# Introduction
The post-glacial relative sea-level fall that characterises most areas of Fennoscandia is fundamental to its archaeology. This follows not only from the dramatic changes to the landscape that this process represent, but also from the fact that if archaeological phenomena were situated on or close to the contemporary shoreline when they were in use, a reconstruction of the trajectory of shoreline displacement can be used to date these phenomena based on their altitude relative to the present day sea-level.

The close association between Norwegian Stone Age sites located in coastal regions and prehistoric shorelines was established by the geologist W.C. Brøgger and his son, the archaeologist A.W. Brøgger, in their respective theses of 1905 (see also @maccurdy1906). The method of shoreline dating has been fundamental to Stone Age research in Norway ever since, and is used both independently and to compliment the dating of sites where other sources of temporal data such as typological indicators or radiometric dates are either limited or not available. Given the coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating by means of shoreline displacement is often the only and most precise and accurate method by which one can hope to date the sites. Dating by means of shoreline displacement is consequently fundamental to our understanding of the Norwegian Stone Age. However, this is both because shoreline dating is fundamental to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shore-line. Consequently, adherence or deviation from this pattern has major implications for the socio-economic foundations of the societies in question.

@@

Despite the fundamental role of shoreline-dating for Norwegian archaeology, there have only been conducted relatively small scale investigations into the applicability of dating by means of shore-line displacement, and the relationship has only been evaluated using relatively coarse methods. The aim of this paper is to provide a systematic and comprehensive review of the degree to which radiocarbon dates correspond with the dates informed by our current knowledge of shoreline displacement in a larger area of south-eastern Norway, using a more refined methodological approach. The goal is to quantify the degree to which the assumption of shore-bound settlement holds through the Stone Age. This can be used both to inform the certainty with which one can depend on dating by means of shoreline displacement, as well as its precision.  If there is chronological variation in the degree to which the sites were shore-bound, the investigation can also help identify societal shifts, and any such variation can potentially also have implications for how archaeological investigations should be directed at different altitudes in the area of analysis. As presented in more detail below, this problem involves the combined evaluation of three major analytical dimensions. One is the questions of when the sites were in use, the second pertains to the reconstruction of the contemporaneous sea-level, and the third follows from the fact that the relation between site and shore-line is inherently a spatial question. Taking inspiration from recent studies that have integrated multiple sources of spatio-temporal uncertainty, a Monte Carlo simulation-based approach was adopted for the analysis.

>>>>

# Introduction
The post-glacial relative sea-level fall that characterises most areas of Fennoscandia is fundamental to its archaeology. This follows not only from the dramatic changes to the landscape that this process has represented throughout history, but also from the fact that if archaeological phenomena were situated on or close to the contemporary shoreline when they were in use, a reconstruction of the trajectory of shoreline displacement can be used to date these phenomena based on their altitude relative to the present day sea-level. This method has long history of use in the region, as is frequently applied within both the research and heritage management sector to date diverse archaeological phenomena such as rock art, harbour constructions and, as is the focus of this study, Stone Age sites,

The close association between Norwegian Stone Age sites located in coastal regions and prehistoric shorelines was established by the geologist W.C. Brøgger and his son, the archaeologist A.W. Brøgger, in their respective theses of 1905 (see also @maccurdy1906). The method of shoreline dating has been fundamental to Stone Age research in Norway ever since. The method is used both independently and to compliment the dating of sites where other sources of temporal data such as typological indicators or radiometric dates are either limited or not available. Given the coarse and fuzzy resolution of established typological frameworks, the vast amount of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating with reference to shoreline displacement is often the only and most precise and accurate method by which one can hope to date the sites. Shoreline dating is consequently fundamental to our understanding of the Norwegian Stone Age, both because it is fundamental to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shore-line. Consequently, adherence or deviation from this pattern also has major implications for the socio-economic foundations of the societies in question.

@@

Despite the fundamental role of shoreline dating for Norwegian archaeology, there have only been conducted relatively small scale investigations into the applicability of dating by means of shore-line displacement, and the relationship has only been evaluated using relatively coarse methods. The aim of this paper is to provide a systematic and comprehensive review of the degree to which radiocarbon dates correspond with the dates informed by our current knowledge of shoreline displacement in a larger area of south-eastern Norway, using a more refined methodological approach. The goal is to quantify the degree to which the assumption of shore-bound settlement holds through the Stone Age. This can be used both to inform the certainty with which one can depend on dating with reference to shoreline displacement, as well as its precision.  If there is chronological variation in the degree to which the sites were shore-bound, the investigation can also help identify societal shifts, and any such variation can potentially also have implications for how archaeological investigations should be directed at different altitudes in the area of analysis. As presented in more detail below, this problem involves the combined evaluation of three major analytical dimensions. One is the questions of when the sites were in use, the second pertains to the reconstruction of the contemporaneous sea-level, and the third follows from the fact that the relation between site and shore-line is inherently a spatial question. Taking inspiration from recent studies that have integrated multiple sources of spatio-temporal uncertainty through Monte Carlo simulation, a similar approach is adopted here.
